---
title: "BDS Toolbox: DV Assignment 1"
author: "Robert Rotaru"
date: "`r Sys.Date()`"
output: html_document
---

<!--
HOW TO KNIT THIS FILE

1. Save this file in the folder you used for the worksheets, next to the
   `data/` folder that holds `get-data.R`. Keep it separate from your
   group's project: your individual work should never end up in your
   group's GitHub repository.
2. Click the `Knit` button at the top of this file (or press
   Ctrl/Cmd + Shift + K). An .html file with the same name appears next to
   this .Rmd.
3. Knitting starts from a clean R session: it only knows what is created
   *in this file*. The setup chunk below loads the packages, and sources
   data/get-data.R to load the data. The first knit downloads the data, so
   you need an internet connection.

If knitting fails:
- "there is no package called ...": install it once in the Console with
  install.packages("...") - not in this file.
- "could not find function": add library(...) for its package to the
  setup chunk.
- "object not found": you created it in the Console, not in this file.
- "cannot open file": check that this file is next to the data/ folder.
- Check that all code chunks have unique labels.
- Google the error message you receive, as it often provides clues about
  what went wrong.

This comment does not show up in the knitted .html, so you can leave it here.
-->

Use this template to help you get started with your assignment. You can add more code chunks as needed. Submit both documents (.Rmd and .html) following the instructions on the assignment.

**If you make more than one visualization per part, please indicate clearly which visualization we should look at when grading**. 

## Setup environment
```{r setup}
knitr::opts_chunk$set(echo = TRUE) # set default echo = TRUE for all code blocks
rm(list = ls()) # Remove any existing objects in memory

# packages - install these once in the Console, not here:
# install.packages(c("tidyverse", "cowplot", "readODS", "plotly", "lintr"))
library(tidyverse)
library(cowplot)
library(readODS)
library(plotly)

# downloads the data into data/raw/ if you don't already have it, then loads
# eindscores, referentieniveaus, schooladviezen and schoolweging
source("data/get-data.R")
```

## Part A: Replicate this plot
### Prepare data

```{r data-prep-1}

school_provider <- eindscores |>
  select(INSTELLINGSCODE, matches("_(AANTAL|GEM)$")) |>
  pivot_longer(
    cols = matches("_(AANTAL|GEM)$"),
    names_to = c("provider", ".value"),
    names_pattern = "(.+)_(AANTAL|GEM)"
  ) |>
  # DUO writes "<5" for small counts -> NA
  mutate(AANTAL = as.numeric(AANTAL)) |>
  filter(!is.na(AANTAL), AANTAL > 0, !is.na(GEM), GEM > 0) |>
  group_by(INSTELLINGSCODE, provider) |>
  summarise(AANTAL = sum(AANTAL), .groups = "drop")

# most schools only use one provider - take whichever tested the most students
primary_provider <- school_provider |>
  group_by(INSTELLINGSCODE) |>
  slice_max(AANTAL, n = 1, with_ties = FALSE) |>
  ungroup() |>
  select(INSTELLINGSCODE, provider)

```



### Create plot 
```{r plot-1}
```


## Part B: Interactive visualizations with plotly
### Prepare data

```{r data-prep-2}
```

### Create interactive plot

```{r plot-2}
```
